Supermartingales and Termination Analysis of Statistical PCF

### 2020-08-05 probabilisticProgrammingMartingales.tex
An example. Some comments. Ongoing.

\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{verbatim}
\newcommand{\tY}{\textsf{Y}}
\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
\newcommand{\tsample}{\textsf{sample}}
\newcommand{\tscore}{\textsf{score}}
\DeclareMathOperator{\red}{red}
\DeclareMathOperator{\nnext}{next}

@@

A position \akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
\end{align*}
The \changed[lo]{\emph{subterm of $M$ at a position $P$}, denoted $M \mid P$,} is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; P & = M \mid P \\
M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
\tY M \mid \tY ; P & = M \mid P \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
\tscore(M) \mid \tscore ; P & = M \mid P
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). A position such that $M\mid P$ does exist is said to occur in $M$. Substitution at a given position, $M[N/P]$, is defined similarly.

Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways. If $P_1$ is an initial segment (i.e.~prefix) of $P_2$ (written as $P_1 \leq P_2$), then $N_2$ is also a subterm of $N_1$. If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint. This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected.

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
%\lo{On first reading, I struggled to parse the following clauses.}
\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occuring at position $P$ in the term $M$.}
\begin{align*}
  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
\end{align*}
In each of these cases, $M \mid P$ is the redex, and reduction takes place at $P$.
\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}

@@

Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
\begin{align*}
& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(M,P)}/P]$ and $\lambda$ does not occur after $@_2$ of $\tY$ in $P$}
\end{align*}

>>>>

\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{verbatim}
\newcommand{\tY}{\textsf{Y}}
%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
\newcommand{\tif}[3]{\textsf{if}\,(#1, #2, #3)} % space-saving concrete syntax
\newcommand{\tsample}{\textsf{sample}}
\newcommand{\tscore}{\textsf{score}}
\DeclareMathOperator{\red}{red}
\DeclareMathOperator{\nnext}{next}

@@

A position \akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
\end{align*}
The \changed[lo]{\emph{subterm of $M$ at a position $P$}, denoted $M \mid P$,} is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; P & = M \mid P \\
M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
\tY M \mid \tY ; P & = M \mid P \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
\tscore(M) \mid \tscore ; P & = M \mid P
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). 
A position such that $M\mid P$ does exist is said to occur in $M$. 
\changed[lo]{Substitution of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
Take, for example, $M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})$ and $P =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2$, then 
\[
M[\tsample / P] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\]}

Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $P_1$ and $P_2$, can overlap in a few different ways. 
If $P_1$ is an initial segment (i.e.~prefix) of $P_2$ (written as $P_1 \leq P_2$), then $N_2$ is also a subterm of $N_1$. If neither $P_1 \leq P_2$ nor $P_1 \geq P_2$, the positions are said to be disjoint. 
This is mostly relevant in that in any substitution at some position, the subterms at all other positions are unaffected. \lo{The preceding sentence is unclear. Typo? Mostly $\mapsto$ most?}

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
%\lo{On first reading, I struggled to parse the following clauses.}
\changed[lo]{The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $P$ in the term $M$.}
\begin{align*}
  \text{if } M \mid P = (\lambda x.s) v,\ & M \to M[s[v/x]/P] \\
  \text{if } M \mid P = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/P] \\
  \text{if } M \mid P = \tY \lambda x. s,\ & M \to M[\lambda z. s[(\tY \lambda x. s)/x] z/P] \text{ where $z$ is not free in $s$}\\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_1/P] \text{ where }r < 0 \\
  \text{if } M \mid P = \tif{\underline r < 0}{s_1}{s_2},\ & M \to M[s_2/P] \text{ where }r \geq 0 \\
  \text{if } M \mid P = \tsample & \text{ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$},\\ & M \to M[\underline r/P] \text{ where } r \in [0,1] \\
  \text{if } M \mid P = \tscore(\underline r),\ & M \to M[\underline r/P].
\end{align*}
In each of these cases, $M \mid P$ is the redex, and reduction takes place at $P$.
%\lo{The definition above is clear. But as a general rule, it is good to illustrate definitions -- especially if they are genuinely new -- with an example or two.}

@@

Unlike in the purely call-by-value case, the version of the reduction relation that takes into account samples is still a general relation rather than a function, so it is denoted ``$\Rightarrow$'' instead of ``$\red$'', and it relates $\biguplus_{M \in \Lambda_0} I^{L_s(M)}$ to itself.
\begin{align*}
& (M,s) \Rightarrow (N,s \circ i(M \to N)) \text{ if either} \\
& \qquad \text{$M \to N$ and the redex is not $\tsample$, or} \\
& \qquad \text{$M \mid P = \tsample$, $N = M[\underline{s(M,P)}/P]$ and $\lambda$ does not occur after $@_2$ or $\tY$ in $P$}
\end{align*}
